$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-12-03 paper/appendix.tex
a pass on technical parts
Start working on the CPP reviews

%% Appendix
\section{Definitions}
\label{app:definitions}

@@

\begin{proof}[Proof (Sketch)]
  By induction on the derivation $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  \begin{itemize}
    \item
      For constructor $\PO$ we get $\Gamma_i \equiv \Psi_i$.
      From $\Delta_i \equiv \lz$ follows that $m \equiv n$.
      Therefore $\Gamma_j \equiv \Psi_j$ and $\constr{end}$ can be applied.

>>>>

%% Appendix

\section{From names to de Bruijn indices and back}
\label{from_to_deBruijn}

\begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\Raw : \Set}
    \; \textsc{Raw}
  \end{mathpar}

  \begin{equation*}
    \begin{aligned}
      \Raw ::= &\; \PO              &&\text{(inaction)}    \\ 
      |& \; (\new{\Name}) \; \Raw         &&\text{(restriction)} \\ 
      |& \; \comp{\Raw}{\Raw}       &&\text{(parallel)}    \\ 
      |& \; \recv{\Name}{\Name}\Raw &&\text{(input)}       \\ 
      |& \; \send{\Name}{\Name}\Raw &&\text{(output)}      \\
    \end{aligned}
  \end{equation*}
  \caption{Grammar with names.}
  \label{fig:syntax-names}
\end{figure}

The syntax of the \picalc{} \cite{Sangio01} using channel names is given by the $\Raw$ grammar in \autoref{fig:syntax-names}.

Channel names and variables range over $x,y,z$ in $\Name$ and processes over $P,Q,R$ in $\Raw$.
Process $\PO$ denotes the terminated process, where no further communications can occur.
Process $(\new{}x)\; P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recv{x}{y} P$ and $\send{x}{y} P$ denote respectively, the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.
Scope restriction $(\new{}x) \; P$ and input $\recv{x}{y} \; P$ are \emph{binders}, they are the only constructs that introduce bound names --- $x$ and $y$ in $P$, respectively.

In order to demonstrate the correspondence between an \picalc{} that uses names and one that uses de Bruijn indices, we provide conversion functions in both directions and prove that they are inverses of each other up to $\alpha$-conversion.

\paragraph*{From names to de Bruijn indices}
When we translate into de Bruijn indices we keep the original binder names around --- they will serve as name hints for when we translate back.
The translation function $\func{fromRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
Named references within the process are substituted with their corresponding de Bruijn index.
We demand that the original process is well-scoped: that all its free variable names appear in $ctx$ --- this is decidable and we therefore automate the construction of such a proof term.
\begin{alignat*}{2}
    &\func{fromRaw} && : (ctx : \Names_n) \; (P : \Raw) \\
    &               && \to \WellScoped \; ctx \; P \to \Process_n
\end{alignat*}

\paragraph*{From de Bruijn indices to names}
The translation function $\func{toRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
As some widely-used languages do, this translation function produces unique variable names.
These unique variable names use the naming scheme $<name-hint>^{<n>}$, where $^{<n>}$ denotes that the name $<name-hint>$ has already been bound $n$ times before.
\begin{equation*}
    \func{toRaw} : (ctx : \Names_n) \to \Process_n \to \Raw
\end{equation*}

\begin{example}[$\func{fromRaw}$ and $\func{toRaw}$]
  \begin{figure}[t]
    \begin{alignat*}{7}
      &P = (\new{x} ) && (\comp {\recv{x}{x} && \send{x}{z} && \PO} {(\new{y}) && (\send{x}{y} && \recv{y}{y} && \PO)}) \\
      &Q = \new{} && (\comp {\recv{0}{} && \send{0}{2} && \PO} {\new{} && (\send{1}{0} && \recv{0}{} && \PO)}) \\
      &R = (\new{x^0} ) && (\comp {\recv{x^0}{x^1} && \send{x^1}{z^0} && \PO} {(\new{y^0}) && (\send{x^0}{y^0} && \recv{y^0}{y^1} && \PO)})
    \end{alignat*}
    \caption{From names to de Bruijn indices and back}
    \label{fig:conversion}
    \end{figure}
  We illustrate the conversion functions from names to de Bruijn indices $(\func{fromRaw}$) and back ($\func{toRaw}$) with three processes $P,Q,R$  in \autoref{fig:conversion}.

  Process $P$ uses names $x,y,z$ and is translated via the conversion function $\func{fromRaw}$ into process $Q$, which uses de Bruijn indices.
  Process $Q$ is then translated via $\func{toRaw}$ into process $R$, which follows the \emph{Barendregt convention}\footnote{The Barendregt variable convention states that all bound variables/names in a process are distinct among each other and from the free variables/names.} and is $\alpha$-equivalent to the original process $P$.
\end{example}

In the following we present the main results that our conversion functions satisfy.

\begin{nilemma}
  Translating from de Bruijn indices to names via $\func{toRaw}$ results in a well-scoped process.
\end{nilemma}

\begin{nilemma}
  Translating from de Bruijn indices to names via $\func{toRaw}$ results in a process that follows the \emph{Barendregt convention}.
\end{nilemma}

\begin{nilemma}
  Translating from de Bruijn indices to names and back via $\func{fromRaw} \circ \func{toRaw}$ results in the same process modulo internal variable name hints.
\end{nilemma}

\begin{nilemma}
  Translating from names to de Bruijn indices and back via $\func{toRaw} \circ \func{fromRaw}$ results in the same process modulo $\alpha$-conversion.
\end{nilemma}

\begin{proof}[Proof]
  All the above results are proved by induction on \textsc{Process} and \textsc{Var} (\autoref{fig:syntax}).
  Complete details can be found in our mechanisation in Agda in \cite{Zalakain2020Agda}.
\end{proof}

\section{Definitions}
\label{app:definitions}

@@

\begin{proof}[Proof]
  By induction on the derivation $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  \begin{itemize}
    \item
      For constructor $\PO$ we get $\Gamma_i \equiv \Psi_i$.
      From $\Delta_i \equiv \lz$ follows that $m \equiv n$.
      Therefore $\Gamma_j \equiv \Psi_j$ and $\constr{end}$ can be applied.

### 2020-12-03 paper/paper.tex
Start working on the CPP reviews

\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{amsmath}
\usepackage{mathbbol}
\usepackage{mathtools}
\usepackage{hyperref}

@@

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first class values, and can therefore be sent and received.
The \picalc{} is at the basis of the design and implementation of programming languages for concurrency, such as Pict \cite{PierceT00} and more recently Go \footnote{https://golang.org}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties, such as deadlock freedom, livelock freedom or termination, and most notably, session types.
These make the \picalc{} a fully-fledged tool for modelling and verifying concurrent and distributed systems.

@@

Linearity influenced the $\pi$-calculus as well.
Kobayashi et al. \cite{KPT96} introduced a typed version of the \picalc{} where the linear type system restricts the usage of channels to exactly once.
%
Meanwhile, linearity acquired a different flavour with the rise of \emph{session types} \cite{H93,THK94,HVK98}, a formalism used in communication-centric programming where linearity ensures that a channel is owned by exactly one communicating participant --- the channel itself may be used multiple times, as specified by its session type.
Linearity in session types is the key ingredient that guarantees communication safety and session fidelity.

@@

In this paper we present \emph{the first full mechanisation in Agda of a \picalc{} with linear, graded and shared types, all under a common unified framework.}

While linear types are useful for resource-aware programming, graded and shared types are useful for a more flexible and 
mainstream programming.

We define a specification for an algebra on \emph{multiplicities} --- how many times a channel can be used --- and use it to apply \emph{leftover typing} for the first time to the \picalc{}, along the lines of Allais \cite{Allais2018a}.
The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
This allows for linear, graded and shared types to be mixed in the same type system.
Ultimately, we can exploit this formalisation as a unified framework for type-safe distributed modelling and programming using a variety of type systems, and build on top of it more advanced types, theories and languages.

@@

\subsection{Contributions}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: we use type level {de Bruijn indices} \cite{deBruijn1972, Dybjer1994} to introduce a syntax of \picalc{} processes that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
   For convenience and readability, the user can convert from and to a syntax using channel names (\autoref{from_to_deBruijn}).
   These conversion functions are shown to be inverses of one another up to $\alpha$-conversion.

\begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\Raw : \Set}
    \; \textsc{Raw}
  \end{mathpar}

  \begin{equation*}
    \begin{aligned}
      \Raw ::= &\; \PO              &&\text{(inaction)}    \\ 
      |& \; (\new{\Name}) \; \Raw         &&\text{(restriction)} \\ 
      |& \; \comp{\Raw}{\Raw}       &&\text{(parallel)}    \\ 
      |& \; \recv{\Name}{\Name}\Raw &&\text{(input)}       \\ 
      |& \; \send{\Name}{\Name}\Raw &&\text{(output)}      \\
    \end{aligned}
  \end{equation*}
  \caption{Grammar with names.}
  \label{fig:syntax-names}
\end{figure}

The syntax of the \picalc{} \cite{Sangio01} using channel names is given by the $\Raw$ grammar in \autoref{fig:syntax-names}.

Channel names and variables range over $x,y,z$ in $\Name$ and processes over $P,Q,R$ in $\Raw$.
Process $\PO$ denotes the terminated process, where no further communications can occur.
Process $(\new{}x)\; P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recv{x}{y} P$ and $\send{x}{y} P$ denote respectively, the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.
Scope restriction $(\new{}x) \; P$ and input $\recv{x}{y} \; P$ are \emph{binders}, they are the only constructs that introduce bound names --- $x$ and $y$ in $P$, respectively.

\subsection{From names to de Bruijn indices and back}
\label{from_to_deBruijn}
In order to demonstrate the correspondence between an \picalc{} that uses names and one that uses de Bruijn indices, we provide conversion functions in both directions and prove that they are inverses of each other up to $\alpha$-conversion.

\paragraph*{From names to de Bruijn indices}
When we translate into de Bruijn indices we keep the original binder names around --- they will serve as name hints for when we translate back.
The translation function $\func{fromRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
Named references within the process are substituted with their corresponding de Bruijn index.
We demand that the original process is well-scoped: that all its free variable names appear in $ctx$ --- this is decidable and we therefore automate the construction of such a proof term.
\begin{alignat*}{2}
    &\func{fromRaw} && : (ctx : \Names_n) \; (P : \Raw) \\
    &               && \to \WellScoped \; ctx \; P \to \Process_n
\end{alignat*}

\paragraph*{From de Bruijn indices to names}
The translation function $\func{toRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
As some widely-used languages do, this translation function produces unique variable names.
These unique variable names use the naming scheme $<name-hint>^{<n>}$, where $^{<n>}$ denotes that the name $<name-hint>$ has already been bound $n$ times before.
\begin{equation*}
    \func{toRaw} : (ctx : \Names_n) \to \Process_n \to \Raw
\end{equation*}

\begin{example}[$\func{fromRaw}$ and $\func{toRaw}$]
  \begin{figure}[t]
    \begin{alignat*}{7}
      &P = (\new{x} ) && (\comp {\recv{x}{x} && \send{x}{z} && \PO} {(\new{y}) && (\send{x}{y} && \recv{y}{y} && \PO)}) \\
      &Q = \new{} && (\comp {\recv{0}{} && \send{0}{2} && \PO} {\new{} && (\send{1}{0} && \recv{0}{} && \PO)}) \\
      &R = (\new{x^0} ) && (\comp {\recv{x^0}{x^1} && \send{x^1}{z^0} && \PO} {(\new{y^0}) && (\send{x^0}{y^0} && \recv{y^0}{y^1} && \PO)})
    \end{alignat*}
    \caption{From names to de Bruijn indices and back}
    \label{fig:conversion}
    \end{figure}
  We illustrate the conversion functions from names to de Bruijn indices $(\func{fromRaw}$) and back ($\func{toRaw}$) with three processes $P,Q,R$  in \autoref{fig:conversion}.

  Process $P$ uses names $x,y,z$ and is translated via the conversion function $\func{fromRaw}$ into process $Q$, which uses de Bruijn indices.
  Process $Q$ is then translated via $\func{toRaw}$ into process $R$, which follows the \emph{Barendregt convention}\footnote{The Barendregt variable convention states that all bound variables/names in a process are distinct among each other and from the free variables/names.} and is $\alpha$-equivalent to the original process $P$.
\end{example}

In the following we present the main results that our conversion functions satisfy.

\begin{nilemma}
  Translating from de Bruijn indices to names via $\func{toRaw}$ results in a well-scoped process.
\end{nilemma}

\begin{nilemma}
  Translating from de Bruijn indices to names via $\func{toRaw}$ results in a process that follows the \emph{Barendregt convention}.
\end{nilemma}

\begin{nilemma}
  Translating from de Bruijn indices to names and back via $\func{fromRaw} \circ \func{toRaw}$ results in the same process modulo internal variable name hints.
\end{nilemma}

\begin{nilemma}
  Translating from names to de Bruijn indices and back via $\func{toRaw} \circ \func{fromRaw}$ results in the same process modulo $\alpha$-conversion.
\end{nilemma}

\begin{proof}[Proof]
  All the above results are proved by induction on \textsc{Process} and \textsc{Var} (\autoref{fig:syntax}).
  Complete details can be found in our mechanisation in Agda in \cite{Zalakain2020Agda}.
\end{proof}

@@

The work by Goto et al. \cite{Goto2016a} is, to the best of our knowledge, the first formalisation of session types which comes along with a mechanised proof of type safety in Coq.
The authors extend session types with polymorphism and pattern matching.
They use a locally-nameless encoding for variable references, a syntax prior to types, and an LTS semantics that encodes session-typed processes into the \picalc{}.
Their type system uses reordering of contexts and extrinsic context splits, whether they are not needed in our work.

@@

In her master's thesis, Zalakain uses a HOAS to embed a session-typed \picalc{} in Coq \cite{Zalakain2019}.
Channel type polymorphism (preventing channels from being constructed out of thin air) and a process-traversing predicate ensure that channels are used linearly --- subject reduction is then shown to preserve this predicate.

@@

Castro et al. \cite{Castro2020} provide tooling for locally-nameless representations of process calculi in Coq  --- in Coq de Bruijn indices are not as popular as in Agda or Idris because dependent pattern matching is far from straightforward.
As a use-case, they use their tool to help automate proofs of subject reduction for a type system with session types.

@@

In this paper we provide a well-scoped syntax and a semantics for the \picalc{}, extrinsically define a type system on top of the syntax capable of handling linear, graded and shared types under the same unified framework and show that reduction preserves the well-typedness of processes.
We avoid the need for extrinsic context splits by defining a type system based on leftover typing \cite{Allais2018a}, which is defined here for the first time for the \picalc{}.
As a result, theorems like framing, weakening and strengthening can now we stated for the linear \picalc{}.
Our work is fully mechanised in around 1850 lines of code in Agda \cite{Zalakain2020Agda}.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{amsmath}
\usepackage{mathbbol}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{todonotes}

@@

The \picalc{} \cite{MilnerPW92,Milner99} is a computational model for communication and concurrency that boils concurrent processing down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first class values, and can therefore be sent and received.
The \picalc{} is at the basis of the design and implementation of programming languages for concurrency, such as Pict \cite{PierceT00} and more recently Go \footnote{https://golang.org}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties, such as deadlock freedom, livelock freedom or termination, and most notably, session types.
These make the \picalc{} a fully-fledged tool for modelling and verifying concurrent and distributed systems.

@@

\todo{introduce shared and graded types sooner}
\todo{too much on session types}
Linearity influenced the $\pi$-calculus as well.
Kobayashi et al. \cite{KPT96} introduced a typed version of the \picalc{} where the linear type system restricts the usage of channels to exactly once.
%
Meanwhile, linearity acquired a different flavour with the rise of \emph{session types} \cite{H93,THK94,HVK98}, a formalism used in communication-centric programming where linearity ensures that a channel is owned by exactly one communicating participant --- the channel itself may be used multiple times, as specified by its session type.
Linearity in session types is the key ingredient that guarantees communication safety and session fidelity.

@@

In this paper we make the following contributions:

\todo{complete}
\begin{enumerate}
  \item We use \textbf{type-level {de Bruijn indices} \cite{deBruijn1972, Dybjer1994} to introduce a syntax for \picalc{} processes that is well-scoped by construction}: every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).

  \item We \textbf{abstract our type system over a set of resource-aware usage algebras}:
    The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
    This allows for linear, graded and shared types to be mixed in the same type system.
    While linear types are useful for resource-aware programming, graded and shared types are useful for a more flexible and mainstream programming.

  \item We \textbf{apply leftover typing to the \picalc{} and prove type preservation for the resulting type system} along the lines of Allais \cite{Allais2018a}.
\end{enumerate}

\subsection{Summary}
\todo{shorten}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: we use type level {de Bruijn indices} \cite{deBruijn1972, Dybjer1994} to introduce a syntax of \picalc{} processes that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
   For convenience and readability, the user can convert from and to a syntax using channel names (\autoref{from_to_deBruijn}).
   These conversion functions are shown to be inverses of one another up to $\alpha$-conversion.

\todo{adapt to de bruijn}
Process $\PO$ denotes the terminated process, where no further communications can occur.
Process $\new{}\; P$ creates a new channel bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recv{x}{} P$ and $\send{x}{y} P$ denote respectively, the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.
Scope restriction $(\new{}x) \; P$ and input $\recv{x}{y} \; P$ are \emph{binders}, they are the only constructs that introduce bound names --- $x$ and $y$ in $P$, respectively.

@@

@@

The work by Goto et al. \cite{Goto2016a} is, to the best of our knowledge, the first formalisation of session types which comes along with a mechanised proof of type safety in Coq.
The authors extend session types with polymorphism and pattern matching.
They use a locally-nameless encoding for variable references, a syntax prior to types, and an LTS semantics that encodes session-typed processes into the \picalc{}.
Their type system uses reordering of contexts and extrinsic context splits, which are not needed in our work.

@@

@@

Castro et al. \cite{Castro2020} provide tooling for locally-nameless representations of process calculi in Coq, where de Bruijn indices are not as popular as in Agda or Idris.
As a use-case, they use their tool to help automate proofs of subject reduction for a type system with session types.

@@

In this paper we provide a well-scoped syntax and a semantics for the \picalc{}, extrinsically define a type system on top of the syntax capable of handling linear, graded and shared types under the same unified framework and show that reduction preserves the well-typedness of processes.
We avoid the need for extrinsic context splits by defining a type system based on leftover typing \cite{Allais2018a}, which is defined here for the first time for the \picalc{}.
As a result, theorems like framing, weakening and strengthening can now be stated for the linear \picalc{}.
Our work is fully mechanised in around 1850 lines of code in Agda \cite{Zalakain2020Agda}.
